Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-10 paper/preprint/paper.tex

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note}]

@@

We find that CT is associated with little to no decrease in average
predictive performance for our synthetic datasets: test accuracy and
F1-scores decrease by at most \textasciitilde1 percentage point, but
generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
Table~\ref{tbl-acc-train}).

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note}, rightrule=.15mm, titlerule=0mm]

@@

We find that CT is associated with little to no decrease in average
predictive performance for our synthetic datasets: test accuracy and
F1-scores decrease by at most \textasciitilde1 percentage point, but
generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
Table~\ref{tbl-acc-train}). Variation across hyperparameters is
negligible as indicated by small standard deviations for these metrics
across the board.

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
